---
title: "Homework for Class 6 - Writing R Functions"
author: 'Jessica Le (PID: A17321021)'
date: "2025-01-24"
output: pdf_document
---

## Can you improve this code analysis? 
```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

> The generated function takes the input of a protein code from the Protein Data Bank to generate a plot of the beta factor of the atoms in chain A of the specified protein. It will work with any set of input protein structures if `plot_pdb` is called. 

```{r}
library(bio3d)
plot_pdb <- function(pdb_code) {
  s <- read.pdb(pdb_code)
  s.chainA <- trim.pdb(s, chain="A", elety="CA")
  s.b <- s.chainA$atom$b
  plotb3(s.b, sse=s.chainA, typ="l", ylab="Bfactor") 
}
```

> This is a breakdown of the generated function. 

```{r}
#Call installed bio3d database
library(bio3d)

#Takes the input of a pdb code and creates pdb file that is read using `read.pdb` 
plot_pdb <- function(pdb_code) {
  s <- read.pdb(pdb_code)

#Reduces the PDB structures to a smaller subset of atoms using function `trim.pdb` and the input is "s"
  s.chainA <- trim.pdb(s, chain="A", elety="CA")
  
#Selecting beta factor from atoms and the input is "s.chainA"
  s.b <- s.chainA$atom$b
  
#Plotting the beta factor vs. residues for the specified atom where the input is "s.b"
  plotb3(s.b, sse=s.chainA, typ="l", ylab="Bfactor") 
}
```

> Running the generated function by calling `plot_pdb()`, the output plots for the three specified proteins are. 

```{r}
sapply(c("4AKE", "1AKE", "1E4Y"), FUN=plot_pdb)
```
```

