---
title: "Lab 9 Tue 2/4: Halloween Candy project"
author: "Jessica Le (PID: A17321021)"
format: html
---

Today we will examind data from 538 common Halloween candy. In particular, we will use ggplot, dplyr, and PCA to make sense of this multivariate dataset. 

## Importing candy data

```{r}
candy_file <- "Lab 9 - candy-data.txt"

candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset? 

```{r}
nrow(candy)
```
There are 85 different types of candy in the dataset. 

> Q2. How many fruity candy types are in this dataset? 

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types in the dataset. 

> How many chocolate candy are there in the dataset? 

```{r}
sum(candy$chocolate)
```
There are 27 chocolate candy in the dataset. 

## What is your favorite candy? 

`winpercent` provides the percentage of people who prefer a specific candy type over another randomly chosen candy from the dataset. A higher value indicates a more popular candy. 

> Q3. What is your favorite candy in the dataset and what is it's `winpercent` value? 

```{r}
candy["Reese's Peanut Butter cup", "winpercent"]
```
```{r}
candy["Reese's Peanut Butter cup", ]$winpercent
```

When paired with another type of candy in the dataset, 84% of the time Reese's Peanut Butter cup, which is my favorite candy, will be picked as the favored one. 

> Q4. What is the `winpercent` value for "Kit Kat"? 

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the `winpercent` value for "Tootsie Roll Snack Bars"? 

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

The `skim()` function in the skimr package that helps give an overview of a given dataset. First, the package needs to be installed using `install.packages("")`. 

```{r}
library(skimr)
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?